---
title: "Homework 2"
author: "Caitlin Jagla"
date: "03/08/2022"
output: pdf_document
---

# Set-up

Load the tidyverse library and the 3 datasets regarding dog breeds.

```{r, message=FALSE,warning=FALSE, results='hide'}

library(tidyverse)

breed_rank <- read_csv("breed_rank.csv")
breed_traits <- read_csv("breed_traits.csv")
trait_desc <- read_csv("trait_description.csv")

```

Use 2 functions of your choice to investigate the datasets so you know what we are working with.

```{r}

breed_rank |> dim_desc()
breed_traits |> dim_desc()
trait_desc |> dim_desc()

breed_rank |> head()
breed_traits |> head()
trait_desc |> head()
```


# 1. New variable

Let's say we would like to know which dog breeds increased most in rank from 2013 to 2020.

### A. Create a new variable called `diff_rank` that is the difference in rank between 2013 and 2020. 

```{r}
breed_rank <- breed_rank |> mutate(diff_rank = `2020 Rank` - `2013 Rank`)
```

### B. Show the 10 breeds that gained the most interest from 2013 to 2020.

```{r}
breed_rank |> 
  arrange(desc(diff_rank)) |> 
  head(n=10) |> 
  select(Breed, `2013 Rank`, `2020 Rank`, diff_rank)

```




# 2. Reshape to long form

Begin with the breed_rank dataset and create a long-form dataset where the numeric year is in one column and the numeric rank is in another column. Save only the Breed, year, rank, and diff_rank columns. Save the result into `breed_rank_long` and show it in the report. `breed_rank_long` should have dimensions 1560 x 4.

```{r, warning=FALSE}
breed_rank_long <-  breed_rank |> 
    pivot_longer(cols = `2013 Rank`:`2020 Rank`, 
                 names_to = "year", values_to = "rank") |>
    separate(col = "year", into = "year", sep = " ") |>
    select(Breed, year, rank, diff_rank)

dim(breed_rank_long)
```



# 3. Plot

Use a line graph to see the rank of each breed over time for the 5 breeds that gained the most popularity from 2013 to 2020.


```{r}

top_diff <- breed_rank |> 
            arrange(desc(diff_rank)) |> 
            head() |> 
            pull(Breed)
            


breed_rank_long |>
  filter(Breed %in% top_diff) |>
  ggplot(aes(color = Breed, x = year, y = rank)) +
  geom_line(aes(group = Breed)) +
  labs(title = "Top Popularity-Gaining Dog Breeds")

```




# 4. Merge

### A. Start with the `breed_rank_long` dataset and create a new dataset that preserves only rows for 2013 and 2020. Call this new dataset `dogs`. Show `dogs` in the report.


```{r}
dogs <- breed_rank_long |>
  filter(year == c(2013, 2020))

dogs |> head()
```




### B. Use a `left_join()` with `dogs` on the left and `breed_traits` on the right. Save the resulting dataset into `dogs` and show it in the report.

```{r}
dogs <- left_join(dogs, breed_traits)

head(dogs)
```



### C. Now that rank and breed traits are in the same dataset, create a plot of your choice to show the relationship between `Barking Level` and 2020 ranking. Write a sentence to interprete your plot. Remember that high rank = more popular.


```{r}
dogs |> filter(year == 2020 & `Barking Level`!=0) |> mutate_at("Barking Level", factor) |>
  ggplot() + 
  geom_boxplot(aes(fill = `Barking Level`, x = `Barking Level`, y = rank)) +
  labs(title = "Effect of Barking Level on Breed Popularity in 2020")
```


There does not appear to be any relationship between barking level and breed popularity in 2020.

